High geomagnetic field intensity recorded by anorthosite xenoliths requires a strongly powered late Mesoproterozoic geodynamo

### 2022-01-28 PNAS_manuscript/PNAS_manuscript.tex
Updates from Overleaf

% Please add a significance statement to explain the relevance of your work
\significancestatement{Acquiring high-fidelity observational records of ancient magnetic field intensity from the remanent magnetization of rocks is crucial for constraining the evolution of Earth's core. However, robust estimates of ancient field strengths are often difficult to recover due to alteration or non-ideal rock magnetic behavior. In this study, we use plagioclase cumulates (anorthosite) that formed in the deep crust and were brought to the near surface by magma where they cooled and acquired thermal remanent magnetizations. These anorthosite xenoliths from the North American Midcontinent Rift have experienced minimal alteration since their emplacement and yield high quality paleointensity estimates. In contrast to predictions of a progressively decaying field, these data record a strong virtual dipole moment 1.1 billion years ago. A strong field persisted over at least a 20-million-year interval indicating the existence of appreciable power sources for Earth's dynamo during the late Mesoproterozoic.}

@@

\begin{abstract}
Obtaining estimates of Earth's magnetic field strength in deep time is complicated by non-ideal rock magnetic behavior in many igneous rocks. In this study, we target anorthosite xenoliths that cooled and acquired their magnetization within ca. 1092 Ma shallowly emplaced diabase intrusions of the North American Midcontinent Rift. In contrast to the diabase which fails to provide reliable paleointensity estimates, the anorthosite xenoliths are unusually high-fidelity recorders yielding high-quality, single-slope paleointensity results that are consistent at specimen and site levels. An average value of $\sim$89 ZAm$^2$ for the virtual dipole moment from the anorthosites with the highest site-level values up to $\sim$140 ZAm$^2$ are higher than that of the dipole component of Earth's magnetic field today. Such high intensities recorded by the anorthosite xenoliths requires the existence of a strongly powered geodynamo at the time. Together with previous paleointensity data from other Midcontinent Rift rocks, our new results indicate that a dynamo with strong power sources persisted for more than 20 million years ca. 1.1 Ga. These data are inconsistent with there being a progressive monotonic decay of Earth's dynamo strength through the Proterozoic Eon.
\end{abstract}

@@

Paleomagnetic records from ancient rocks are one of the few types of observational data that have the potential to provide constraints on the thermal evolution of Earth’s core. Evidence for a persistent magnetic field through the Proterozoic, for example, likely necessitates the existence of plate tectonics that sustained core-mantle boundary heat flow \cite{Swanson-Hysell2021c}. However, strikingly low estimates of geomagnetic field strengths have been obtained ca. 565 Ma during the Ediacaran Period \cite{Bono2019a, Shcherbakova2019a,Thallner2021b} and ca. 400 Ma during the Devonian Period \cite{Shcherbakova2017a, Shcherbakova2021a, Hawkins2021a}, potentially indicating unusual periods of core dynamo regime at those times. The Ediacaran data have been interpreted to indicate that there was a progressively decaying field up to that time that was followed soon after by initial crystallization of the inner core \cite{Bono2019a}.

@@

The hypothesis that a weak Ediacaran geomagnetic field is a telltale sign of the lack of an inner core with core nucleation following shortly thereafter may predict that it is the most significant weak to strong field transition in the paleointensity record. However, Fig. \ref{fig:PINT_compilation} shows that transitions from low to high field intensities occurred before, during, and after the Ediacaran. In the Ediacaran record developed to date, there is a two-fold increase in Earth's virtual dipole moment when comparing estimates from the ca. 565 Ma Sept-$\hat{I}$les intrusions \cite{Bono2019a} to those from ca. 550 Ma volcanics of the Skinner Cove Formation \cite{Thallner2021a} (Fig. \ref{fig:PINT_compilation}). In the late Mesoproterozoic, there is at least a six-fold increase within a period of $\sim$40 Myr from a low average virtual dipole moment of $\sim$13 ZAm$^2$ recorded by the ca. 1140 Ma Abitibi dikes which yielded straight Arai plots \cite{Macouin2003a}, to a high moment of $\sim$70 ZAm$^2$ recorded by the ca. 1106 Ma Osler Volcanics, with even stronger values from ca. 1092 Ma by Beaver River anorthosite xenoliths that record virtual dipole moments up to $\sim$140 ZAm$^2$. While the Abitibi dikes paleointensity estimates do not go to values as low as the Sept-$\hat{I}$les intrusions, this virtual dipole moment increase in the Mesoproterozoic is the largest yet documented in the Precambrian (Fig. \ref{fig:PINT_compilation}). The tempo and scale of this field intensity transition could match with model-based predictions associated with the onset of inner core nucleation \cite{Davies2021a}. This timing would be broadly consistent with that proposed by \citealp{Biggin2015a}. However, the model prediction would be of sustained strong field values following inner core nucleation which is challenged by data from the ca. 1070 Ma Bangemall Sills which include a sill with a low virtual dipole moment of $\sim$6.4 ZAm$^2$ (Fig. \ref{fig:PINT_compilation};  \citealp{Lloyd2021b}). Following the Ediacaran, there are also low paleointensity estimates from the ca. 414 Ma Strathmore lava flows \cite{Hawkins2021a}, the ca. 410-380 Ma lava flows of the Minusa Basin and the Kola Peninsula \cite{Shcherbakova2017a}, the ca. 408–393 Ma Buribay volcanic complex \cite{Shcherbakova2021a}, as well as the ca. 370 Ma dikes and lavas of the Siberian Viluy Traps \cite{Hawkins2019a} (Fig. \ref{fig:PINT_compilation}). This interval of time with low paleointensity estimates has been termed the ``Mid-Paleozoic Dipole Low'' \cite{Hawkins2021a}. This  ``Mid-Paleozoic Dipole Low'' is followed by high paleointensity values such that there is at least a four-fold increase from a virtual dipole moment of $\sim$23 ZAm$^2$ at ca. 338 to $\sim$93 ZAm$^2$ at ca. 302 Ma (Fig. \ref{fig:PINT_compilation}).

Numerical models have shown that the stability of the geomagnetic axial dipole moment is sensitive to the total heat flux and the pattern of the heat flow distribution through the core-mantle boundary when there are strong available power sources to the geodynamo \cite{Glatzmaier1999a, Olson2007a, Olson2010a}. Modeling results form \cite{Olson2010a} show that a low total heat flux through the core-mantle boundary can prevent the axial dipole from reversing whereas a high heat flux through the boundary can result in increase in reversal frequency and decrease in the axial dipole intensity. The ``Mid-Paleozoic Dipole Low'' has been hypothesized to be the result of a high core-mantle boundary heat flux condition in a framework that includes available power from a crystallizing inner core \cite{Hawkins2019a}, thereby explaining the observed low paleointensities which are nearly as weak as those of the Sept-$\hat{I}$les intrusions \cite{Bono2019a} (Fig. \ref{fig:PINT_compilation}). There can also be large changes in the strength of the geomagnetic axial dipole due to changes in the geometry of core-mantle boundary heat flux. Such spatial heat flow variations could be associated with changes in the structure of the deep mantle due to plumes \cite{Larson1991a, Courtillot2007a}, subduction slabs \cite{Tan2002b, Hounslow2018a}, and distribution of dense mantle basal piles \cite{Biggin2012a}.

Could core-mantle interactions similar to that today exist before the nucleation of the inner core and modulated the geodynamo? With emerging paleomagnetic evidence supporting the existence of modern-day-like plate tectonics that extends through the Proterozoic \cite{Swanson-Hysell2021c} and to the Archean \cite{Mitchell2014a}, kinetic and thermal energy exchange between deep and shallow Earth likely existed since the early time of Earth's history. If the mantle forcing associated with these activities exists regardless of the present of inner core, it could have contributed to the anomalously low or high paleointensity values observed during the Phanerozoic and Proterozoic (Fig. \ref{fig:PINT_compilation}). The unknown magnitude of such forcing may challenge our ability to detect the increase in surface geomagnetic field strength predicted to have happened at the onset of the crystallization of the inner core \cite{Davies2021a} based on paleointensity data alone.

Regardless, the high-fidelity paleointensity recorders of the Beaver River anorthosite xenoliths in the well-preserved Midcontinent Rift record high paleointensities values and necessitates that there was a strongly powered late Mesoproterozoic geodynamo.

>>>>

% Please add a significance statement to explain the relevance of your work
\significancestatement{Acquiring high-fidelity observational records of ancient magnetic field intensity from the remanent magnetization of rocks is crucial for constraining the long-term evolution of Earth's core. However, robust estimates of ancient field strengths are often difficult to recover due to alteration or non-ideal rock magnetic behavior. In this study, we use plagioclase cumulates (anorthosite) that formed in the deep crust and were brought to the near surface by magma where they cooled and acquired thermal remanent magnetizations. These anorthosite xenoliths have experienced minimal alteration since their emplacement and yield high quality paleointensity estimates. In contrast to predictions of a progressively decaying field leading up to late nucleation of Earth's inner core, these data record a strong virtual dipole moment 1.1 billion years ago. A strong field persisted over at least a 20-million-year interval indicating the existence of appreciable power sources for Earth's dynamo during the late Mesoproterozoic.}

@@

\begin{abstract}
Obtaining estimates of Earth's magnetic field strength in deep time is complicated by non-ideal rock magnetic behavior in many igneous rocks. In this study, we target anorthosite xenoliths that cooled and acquired their magnetization within ca. 1092 Ma shallowly emplaced diabase intrusions of the North American Midcontinent Rift. In contrast to the diabase which fails to provide reliable paleointensity estimates, the anorthosite xenoliths are unusually high-fidelity recorders yielding high-quality, single-slope paleointensity results that are consistent at specimen and site levels. An average value of $\sim$89 ZAm$^2$ for the virtual dipole moment from the anorthosites with the highest site-level values up to $\sim$140 ZAm$^2$ are higher than that of the dipole component of Earth's magnetic field today. Such high intensities recorded by the anorthosite xenoliths requires the existence of a strongly powered geodynamo at the time. Together with previous paleointensity data from other Midcontinent Rift rocks, our new results indicate that a dynamo with strong power sources persisted for more than 20 million years ca. 1.1 Ga. These data are inconsistent with there being a progressive monotonic decay of Earth's dynamo strength through the Proterozoic Eon and could challenge the hypothesis of a young inner core.
\end{abstract}

@@

Paleomagnetic records from ancient rocks are one of the few types of observational data that have the potential to provide constraints on the thermal evolution of Earth’s core. Evidence for a persistent magnetic field through the Proterozoic, for example, likely necessitates the existence of plate tectonics that sustained core-mantle boundary heat flow \cite{Swanson-Hysell2021c}. However, strikingly low estimates of geomagnetic field strengths have been obtained ca. 565 Ma during the Ediacaran Period \cite{Bono2019a, Shcherbakova2019a,Thallner2021b} and ca. 370 Ma during the Devonian Period \cite{Shcherbakova2017a, Shcherbakova2021a, Hawkins2021a}, potentially indicating unusual periods of core dynamo regime at those times. The Ediacaran data have been interpreted to indicate that there was a progressively decaying field up to that time that was followed soon after by initial crystallization of the inner core \cite{Bono2019a}.

@@

The hypothesis that a weak Ediacaran geomagnetic field is a telltale sign of the lack of an inner core with core nucleation following shortly thereafter may predict that it is the most significant weak to strong field transition in the paleointensity record. However, Fig. \ref{fig:PINT_compilation} shows that transitions from low to high field intensities occurred before, during, and after the Ediacaran. In the Ediacaran record developed to date, there is a two-fold increase in Earth's virtual dipole moment when comparing estimates from the ca. 565 Ma Sept-$\hat{I}$les intrusions \cite{Bono2019a} to those from ca. 550 Ma volcanics of the Skinner Cove Formation \cite{Thallner2021a} (Fig. \ref{fig:PINT_compilation}). In the late Mesoproterozoic, there is at least a six-fold increase within a period of $\sim$40 Myr from a low average virtual dipole moment of $\sim$13 ZAm$^2$ recorded by the ca. 1140 Ma Abitibi dikes which yielded straight Arai plots \cite{Macouin2003a}, to a high moment of $\sim$70 ZAm$^2$ recorded by the ca. 1106 Ma Osler Volcanics, with even stronger values from ca. 1092 Ma by Beaver River anorthosite xenoliths that record virtual dipole moments up to $\sim$140 ZAm$^2$. While the Abitibi dikes paleointensity estimates do not go to values as low as the Sept-$\hat{I}$les intrusions, this virtual dipole moment increase in the Mesoproterozoic is the largest yet documented in the Precambrian on a 10s of millions of year timescale (Fig. \ref{fig:PINT_compilation}). The tempo and scale of this field intensity transition could match with model-based predictions associated with the onset of inner core nucleation \cite{Davies2021a}. This timing would be broadly consistent with that proposed by \citealp{Biggin2015a}. However, a model prediction could be of sustained strong field values following inner core nucleation which is challenged by data from the ca. 1070 Ma Bangemall Sills which include a sill with a low virtual dipole moment of $\sim$6.4 ZAm$^2$ (Fig. \ref{fig:PINT_compilation};  \citealp{Lloyd2021b}). Following the Ediacaran, there are also low paleointensity estimates from Devonian rocks such as ca. 370 Ma dikes and lavas of the Siberian Viluy Traps that give virtual dipole moment of 4.3 to 14.9 ZAm$^2$ (Fig. \ref{fig:PINT_compilation}; \citealp{Hawkins2019a}). These low values and as well as data from the ca. 414 Ma Strathmore lava flows \cite{Hawkins2021a}, the ca. 410-380 Ma lava flows of the Minusa Basin and the Kola Peninsula \cite{Shcherbakova2017a}, the ca. 408–393 Ma Buribay volcanics  volcanic complex \cite{Shcherbakova2021a} and the ca. 332 Ma Kinghorn volcanics \cite{Hawkins2021a} has led to the proposal of this interval as the ``Mid-Paleozoic Dipole Low'' \cite{Hawkins2021a}. This  ``Mid-Paleozoic Dipole Low'' is followed by high paleointensity values such that there is a six-fold increase in the database from virtual dipole moment of $\sim$16 ZAm$^2$ at ca. 332 Ma to $\sim$99 ZAm$^2$ ca. 308 Ma in the late Carboniferous \cite{Hawkins2021a}.

Given that not each record of a weak field can be a minimum prior to inner core nucleation, what processes could lead to a weak dipole at Earth's surface even in the presence of a crystallizing inner core? Numerical models have shown that the dipole moment is sensitive to both the magnitude and spatial pattern of heat flow across the core-mantle boundary when there are strong available power sources to the geodynamo \cite{Olson2007a, Olson2010a}. In such models, relatively low total heat flux across the core-mantle boundary can prevent the axial dipole from reversing whereas a high heat flux through the boundary can result in an increase in reversal frequency and decrease in the axial dipole intensity. The ``Mid-Paleozoic Dipole Low'' has been hypothesized to be the result of elevated core-mantle boundary heat flux condition at a time when there was also available power from a crystallizing inner core \cite{Hawkins2019a}, thereby explaining the observed low paleointensities which include values as weak as those of the ca. 565 Ma Sept-$\hat{I}$les intrusions \cite{Bono2019a} (Fig. \ref{fig:PINT_compilation}). Mantle convection can modulate core mantle boundary heat flow through changes in the structure of the deep mantle associated with upwelling plumes \cite{Larson1991a, Courtillot2007a} and subducted slabs \cite{Tan2002b, Biggin2012a, Hounslow2018a}. Strong evidence for differential plate tectonic motion extends back to ca. 2.2 Ga in the Paleoproterozoic \cite{Mitchell2014a,Swanson-Hysell2021b} and potentially back to ca. 3.2 Ga in the Archean \cite{Brenner2020a}. Plate tectonic modulations of core mantle boundary heat flow are therefore expected throughout the Proterozoic. Such changes may explain large variability in values as in the Phanerozoic \cite{Lloyd2021a} and may challenge our ability to detect the increase in surface geomagnetic field strength predicted to have happened at the onset of inner core crystallization.

Overall, the high-fidelity paleointensity recorders of the Beaver River anorthosite xenoliths in the well-preserved Midcontinent Rift record strong field strengths 1.1 billion years ago. The highest site level field strength would rank in the xx percentile of the paleointensity database for the Cenozoic Era when there was unequivocally a crystallizing inner core. These high surface field strengths necessitate appreciable power to the late Mesoproterozoic geodynamo.

% Could core-mantle interactions similar to that today exist before the nucleation of the inner core and modulated the geodynamo? With emerging paleomagnetic evidence supporting the existence of modern-day-like plate tectonics that extends through the Proterozoic \cite{Swanson-Hysell2021c} and to the Archean \cite{Mitchell2014a}, kinetic and thermal energy exchange between deep and shallow Earth likely existed since the early time of Earth's history. If the mantle forcing associated with these activities exists regardless of the present of inner core, it could have contributed to the anomalously low or high paleointensity values observed during the Phanerozoic and Proterozoic (Fig. \ref{fig:PINT_compilation}).
